---
title: "PUBLPOL 218X Assignment 3"
author: "Tiffany Zhu"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    number_sections: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# PUMS Multiple Regression: Marital Status

I have read a few articles [cite sources] that claim that marital status increasingly depends on socioeconomic status. Supposedly, those who get married are increasingly high-income, well-educated individuals, while lower-income and less-educated people are choosing not to marry.

I will test those claims by studying the relationship between marital status (which I set as my dependent variable) and income and educational attainment (which I set as my independent variables). This analysis will be at the individual level. I will focus on individuals aged 18 or older, since individuals under 18 rarely marry.




``` {r data1, echo = T}

```

# CBG/Tract/ZCTA Regression: X, Y, and Z



``` {r data2, echo = T}

```