---
title: "PUBLPOL 218X Assignment 3"
author: "Tiffany Zhu"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    number_sections: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("~/Documents/GitHub/tiffanyzhu10.github.io/assignment3")
library(tidyverse)
library(plotly)
library(sf)
library(tigris)
library(leaflet)
library(zoo)
library(censusapi)
library(tidycensus)

census_api_key("5ace7165d1f551af1a6a8876ee3986faa544ea83")

Sys.setenv(CENSUS_KEY="5ace7165d1f551af1a6a8876ee3986faa544ea83")
```

# PUMS Multiple Regression: Marital Status

I have read a few articles [cite sources] that claim that marital status increasingly depends on socioeconomic status. Supposedly, those who get married are increasingly high-income, well-educated individuals, while lower-income and less-educated people are choosing not to marry.

I will test those claims by studying the relationship between marital status (which I set as my dependent variable) and income and educational attainment (which I set as my independent variables). This analysis will be at the census tract level. I will focus on individuals aged 18 or older, since individuals under 18 rarely marry.

``` {r pums_setup, eval = F}

ca_pums = get_pums(variables = c("PUMA", "AGEP", "MAR", "SCHL", "PINCP"),
                   state = "CA",
                   year = 2019,
                   survey = "acs1",
                   recode = T)
saveRDS(ca_pums, file = "ca_pums.rds")

```

``` {r pums1, echo = T}
# Get PUMS and PUMAS data

pums_vars_2019 = pums_variables %>%
                 filter(year == 2019, survey == "acs1")
pums_vars_2019_inds = pums_vars_2019 %>%
                      distinct(var_code, var_label, data_type, level) %>%
                      filter(level == "person")
ca_pums = readRDS("ca_pums.rds")
ca_pumas = pumas("CA", cb = T, progress_bar = F)

# Isolate Bay Area PUMS

bay_county_names =
  c(
    "Alameda",
    "Contra Costa",
    "Marin",
    "Napa",
    "San Francisco",
    "San Mateo",
    "Santa Clara",
    "Solano",
    "Sonoma"
  )

ca_counties = counties("CA", cb = T, progress_bar = F)
bay_counties = ca_counties %>%
               filter(NAME %in% bay_county_names)

bay_pumas = ca_pumas %>%
            st_centroid() %>%
            .[bay_counties, ] %>%
            mutate(PUMACE10 = as.numeric(PUMACE10)) %>%
            st_set_geometry(NULL) %>%
            left_join(ca_pumas %>% select(GEOID10)) %>%
            st_as_sf()
bay_pums = ca_pums %>%
           mutate(PUMA = as.numeric(PUMA)) %>%
           filter(PUMA %in% bay_pumas$PUMACE10)

bay_pums
```

``` {r pums2, echo = T}
# Clean dataset
# Filter out all children under age 3 (SCHL == bb)
# Convert "character"-encoded variables MAR and SCHL to doubles
# Filter dataset so all individuals are over 18 years old
# Create new variable "married" representing number of ever-married individuals in each PUMA
# Create new variable "college" representing number of college-educated individuals in each PUMA
# Collapse to PUMAs
# Calculate percentage of married and college-educated people in each PUMA

bay_marriage = bay_pums %>%
               filter(SCHL != "bb") %>%
               mutate(MAR = as.numeric(MAR),
                      SCHL = as.numeric(SCHL)) %>%
               filter(AGEP >= 18) %>%
               mutate(married = ifelse(
                        (MAR < 5),
                        PWGTP,
                        0),
                      college = ifelse(
                        (SCHL >= 21),
                        PWGTP,
                        0
                      )
               ) %>%
               group_by(PUMA) %>%
               summarize(mean_personal_income = weighted.mean(PINCP, PWGTP, na.rm = T),
                         perc_married = sum(married, na.rm = T)/
                                        sum(PWGTP, na.rm = T)*100,
                         perc_college = sum(college, na.rm = T)/
                                        sum(PWGTP, na.rm = T)*100)

bay_marriage
```

``` {r pums3, echo = T}
# Regression

model = lm(perc_married ~ mean_personal_income, bay_marriage)
summary(model)

ggplot(data = bay_marriage,
       aes(x = mean_personal_income,
           y = perc_married)) +
  geom_point() +
  geom_smooth(method = "lm")

```

# CBG/Tract/ZCTA Regression: X, Y, and Z



``` {r data2, echo = T}

```