---
title: "PUBLPOL 218X Assignment 3"
author: "Tiffany Zhu"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    number_sections: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("~/Documents/GitHub/tiffanyzhu10.github.io/assignment3")
library(tidyverse)
library(plotly)
library(sf)
library(tigris)
library(leaflet)
library(zoo)
library(censusapi)
library(tidycensus)

census_api_key("5ace7165d1f551af1a6a8876ee3986faa544ea83")

Sys.setenv(CENSUS_KEY="5ace7165d1f551af1a6a8876ee3986faa544ea83")
```

# PUMS Multiple Regression: Marital Status

I have read a few articles [cite sources] that claim that marital status increasingly depends on socioeconomic status. Supposedly, those who get married are increasingly high-income, well-educated individuals, while lower-income and less-educated people are choosing not to marry.

I will test those claims by studying the relationship between marital status (which I set as my dependent variable) and income and educational attainment (which I set as my independent variables). This analysis will be at the individual level. I will focus on individuals aged 18 or older, since individuals under 18 rarely marry.

``` {r pums_setup, eval = F}

ca_pums = get_pums(variables = c("PUMA", "MAR", "SCHL", "PINCP"),
                   state = "CA",
                   year = 2019,
                   survey = "acs1",
                   recode = T)
saveRDS(ca_pums, file = "ca_pums.rds")

```

# CBG/Tract/ZCTA Regression: X, Y, and Z



``` {r data2, echo = T}

```